Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-06-02 aosd13/paperBody.tex
Reformat code in Ch. 5 for wider column

def titleFilterQuery(records: Exp[Set[BookData]], 
                keyword: Exp[String]) = for {
  record <- records
  if record.title.contains(keyword)
} yield (record.title, record.authorName)
val resQuery = titleFilterQuery(recordsQuery, "Principles")
val res = resQuery.optimize.eval
\end{lstlisting}
\caption{\LoS\ version of Fig.~\ref{fig:query}; \code{recordQuery} contains a reification of the query, \code{records} its result.}
\label{fig:reifiedQuery}
\end{figure}

>>>>

def titleFilterQuery(records: Exp[Set[BookData]], keyword: Exp[String]) = for {
  record <- records
  if record.title.contains(keyword)
} yield (record.title, record.authorName)

val resQuery = titleFilterQuery(recordsQuery, "Principles")
val res = resQuery.optimize.eval
\end{lstlisting}
\caption{\LoS\ version of Fig.~\ref{fig:query}; \code{recordQuery} contains a reification of the query, \code{records} its result.}
\label{fig:reifiedQuery}
\end{figure}

### 2017-06-02 aosd13/pepm13/caseStudyImpl.tex
Reformat code in Ch. 5 for wider column

%\pg{Somewhere: Contrast a bit with LMS - we are more open, provide less compile-time safety, etc.; motivate differences. This section risks being boring, explain why it is not.}
\section{Representing expression trees}
\label{subsec:repexptrees}
In the previous section, we have seen that expressions that would have type \code{T} in a native Scala query are reified and have type \code{Exp[T]} in \LoS. The generic type \code{Exp[T]} is the base for our reification of Scala expression as expression trees, that is, as data structures in memory. We provide a subclass of \code{Exp[T]} for every different form of expression we want to reify. For example, in Fig.~\ref{fig:query} the expression \code{author.firstName + " " + author.lastName} must be reified even though it is not collection-related, for otherwise the optimizer could not see whether \code{author} is used. Knowing this is needed for instance to remove variables which are bound but not used. Hence, this expression is reified as \code{StringConcat(StringConcat(AuthorFirstName(author), Const(" ")), AuthorLastName(author))}. This example uses the constructors of the following subclasses of \code{Exp[T]} to create the expression tree.

\begin{lstlisting}
case class Const[T](t: T) extends Exp[T]

case class StringConcat(str1: Exp[String],
                        str2: Exp[String])
  extends Exp[String]

case class AuthorFirstName(t: Exp[Author])
  extends Exp[String]

case class AuthorLastName(t: Exp[Author])
  extends Exp[String]
\end{lstlisting}

@@

To this end, consider again the fragment \code{author.firstName + " " + author.lastName}, now in the context of the {\LoS}-enabled query in Fig.~\ref{fig:reifiedQuery}. It looks like a normal Scala expression, even syntactically unchanged from Fig.~\ref{fig:query}. However, evaluating that code in the context of Fig.~\ref{fig:reifiedQuery} does not concatenate any strings, but creates an expression tree instead. Although the code looks like the same expression, it has a different \emph{type}, \code{Exp[String]} instead of \code{String}. This difference in the type is caused by the context: The variable \code{author} is now bound in a {\LoS}-enabled query and therefore has type \code{Exp[Author]} instead of \code{Author}. We can still access the \code{firstName} field of \code{author}, because expression trees of type \code{Exp[T]} provide the same interface as values of type \code{T}, except that all operations return expressions trees instead of values.

@@

implicit def expToAuthorOps(t: Exp[Author]) =
  new AuthorOps(t)
implicit def toAuthorOps(t: Author) =
  expToAuthorOps(pure(t))
class AuthorOps(t: Exp[Author]) {
  def firstName: Exp[String] = AuthorFirstName(t)
  def lastName: Exp[String] = AuthorLastName(t)
}
\end{lstlisting}

@@

\begin{lstlisting}
implicit def tuple2ToTuple2Exp[A1, A2]
    (tuple: (Exp[A1], Exp[A2])): LiftTuple2[A1, A2] =
  LiftTuple2[A1, A2](tuple._1, tuple._2)
case class LiftTuple2[A1, A2](t1: Exp[A1], t2: Exp[A2]) extends Exp[(A1, A2)]
\end{lstlisting}

@@

\begin{lstlisting}
implicit def expToListOps[T](coll: Exp[List[T]]) = 
  new ListOps(coll)
implicit def toListOps[T](coll: List[T]) = 
  expToListOps(coll)
class ListOps(coll: Exp[List[T]]) {
  def map[U](f: Exp[T] => Exp[U]) =
    ListMapNode(coll, Fun(f))
}
case class ListMapNode[T, U](coll: Exp[List[T]], mapping: Exp[T => U]) extends Exp[List[U]]
\end{lstlisting}

@@

One consequence of this syntactic desugaring is that a single for-comprehension can operate over different collection types%
%\footnote{In Haskell 98, instead, this is not possible.}%
. The type of the result depends essentially on the type of the root collection, that is \code{books} in the example above. The example above can hence be altered to produce a sequence rather than a set by simply converting the root collection to another type:
\begin{lstlisting}
val recordsSeq = for {
   book <- books£\textbf{.toSeq}£
   if book.publisher == "Pearson Education"
   author <- book.authors
  } yield BookData(book.title,
    author.firstName + " " + author.lastName,
    book.authors.size - 1)
\end{lstlisting}

@@

\begin{lstlisting}
def map[U](f: T => U)
  (implicit cbf: CanBuildFrom[Repr, U, That]): That
def flatMap[U](f: T => Traversable[U])
  (implicit cbf: CanBuildFrom[Repr, U, That]): That
def withFilter[U](p: T => Boolean): Repr.
\end{lstlisting}

@@

\smartParagraph{The lifting}
The basic idea is to use the enrich-my-library pattern to lift collection methods from \code{TraversableLike[T, Repr]} to \code{TraversableLikeOps[T, Repr]}:
\begin{lstlisting}
implicit def expToTraversableLikeOps[T, Repr]
  (v: Exp[TraversableLike[T, Repr]]) =
  new TraversableLikeOps[T, Repr](v)
\end{lstlisting}

@@

%We use enrich-my-library again with an implicit conversion to \code{TraversableLikeOps[T, Repr]}, shown later, which corresponds to
%\code{TraversableLike[T, Repr]}.
%We provide an implicit conversion to \code{TraversableLikeOps[T, Repr]} from \code{Exp[Repr]} for any type \code{Repr} which is a subtype of \code{TraversableLike[T, Repr]} and \code{Traversable[T]} (for some element type \code{T}). This can be expressed as:
\begin{lstlisting}
implicit def expToTraversableLikeOps
  [T, Repr <: Traversable[T]
                with TraversableLike[T, Repr]]
  (v: Exp[Repr]) =
    new TraversableLikeOps[T, Repr](v)
\end{lstlisting}

@@

\begin{figure}
\begin{center}
\begin{lstlisting}
class TraversableLikeOps[T, Repr <: Traversable[T]
        with TraversableLike[T, Repr]](t: Exp[Repr]) {
  val t: Exp[Repr]
  def withFilter(f: Exp[T] => Exp[Boolean]): Exp[Repr] =
    Filter(this.t, FuncExp(f))
  def map[U, That <: TraversableLike[U, That]]
         (f: Exp[T] => Exp[U])
         (implicit c: CanBuildFrom[Repr, U, That]):
         Exp[That] =
    MapNode(this.t, FuncExp(f))
  //... other methods ...
}
//definitions of MapNode, Filter omitted.
\end{lstlisting}
\end{center}
\caption{Lifting \code{TraversableLike}}
\label{fig:liftTrav}
\end{figure}

>>>>

%\pg{Somewhere: Contrast a bit with LMS - we are more open, provide less compile-time safety, etc.; motivate differences. This section risks being boring, explain why it is not.}
\section{Representing expression trees}
\label{subsec:repexptrees}
In the previous section, we have seen that expressions that would have type \code{T} in a native Scala query are reified and have type \code{Exp[T]} in \LoS. The generic type \code{Exp[T]} is the base for our reification of Scala expression as expression trees, that is, as data structures in memory. We provide a subclass of \code{Exp[T]} for every different form of expression we want to reify. For example, in Fig.~\ref{fig:query} the expression \code{author.firstName + " " + author.lastName} must be reified even though it is not collection-related, for otherwise the optimizer could not see whether \code{author} is used. Knowing this is needed for instance to remove variables which are bound but not used. Hence, this expression is reified as
\begin{lstlisting}
StringConcat(StringConcat(AuthorFirstName(author), Const(" ")), AuthorLastName(author))
\end{lstlisting}
This example uses the constructors of the following subclasses of \code{Exp[T]} to create the expression tree.

\begin{lstlisting}
case class Const[T](t: T) extends Exp[T]
case class StringConcat(str1: Exp[String], str2: Exp[String]) extends Exp[String]
case class AuthorFirstName(t: Exp[Author]) extends Exp[String]
case class AuthorLastName(t: Exp[Author]) extends Exp[String]
\end{lstlisting}

@@

To this end, consider again the fragment
\begin{lstlisting}
author.firstName + " " + author.lastName
\end{lstlisting}
now in the context of the {\LoS}-enabled query in Fig.~\ref{fig:reifiedQuery}. It looks like a normal Scala expression, even syntactically unchanged from Fig.~\ref{fig:query}. However, evaluating that code in the context of Fig.~\ref{fig:reifiedQuery} does not concatenate any strings, but creates an expression tree instead. Although the code looks like the same expression, it has a different \emph{type}, \code{Exp[String]} instead of \code{String}. This difference in the type is caused by the context: The variable \code{author} is now bound in a {\LoS}-enabled query and therefore has type \code{Exp[Author]} instead of \code{Author}. We can still access the \code{firstName} field of \code{author}, because expression trees of type \code{Exp[T]} provide the same interface as values of type \code{T}, except that all operations return expressions trees instead of values.

@@

implicit def expToAuthorOps(t: Exp[Author]) = new AuthorOps(t)
implicit def toAuthorOps(t: Author) = expToAuthorOps(pure(t))

class AuthorOps(t: Exp[Author]) {
  def firstName: Exp[String] = AuthorFirstName(t)
  def lastName: Exp[String] = AuthorLastName(t)
}
\end{lstlisting}

@@

\begin{lstlisting}
implicit def tuple2ToTuple2Exp[A1, A2](tuple: (Exp[A1], Exp[A2])): LiftTuple2[A1, A2] =
  LiftTuple2[A1, A2](tuple._1, tuple._2)
case class LiftTuple2[A1, A2](t1: Exp[A1], t2: Exp[A2]) extends Exp[(A1, A2)]
\end{lstlisting}

@@

\begin{lstlisting}
implicit def expToListOps[T](coll: Exp[List[T]]) = new ListOps(coll)
implicit def toListOps[T](coll: List[T]) = expToListOps(coll)

class ListOps(coll: Exp[List[T]]) {
  def map[U](f: Exp[T] => Exp[U]) = ListMapNode(coll, Fun(f))
}

case class ListMapNode[T, U](coll: Exp[List[T]], mapping: Exp[T => U]) extends Exp[List[U]]
\end{lstlisting}

@@

One consequence of this syntactic desugaring is that a single for-comprehension can operate over different collection types.
%\footnote{In Haskell 98, instead, this is not possible.}%
The type of the result depends essentially on the type of the root collection, that is \code{books} in the example above. The example above can hence be altered to produce a sequence rather than a set by simply converting the root collection to another type:
\begin{lstlisting}
val recordsSeq = for {
   book <- books£\textbf{.toSeq}£
   if book.publisher == "Pearson Education"
   author <- book.authors
  } yield BookData(book.title, author.firstName + " " + author.lastName,
    book.authors.size - 1)
\end{lstlisting}

@@

\begin{lstlisting}
def map[U](f: T => U)(implicit cbf: CanBuildFrom[Repr, U, That]): That
def flatMap[U](f: T => Traversable[U])(implicit cbf: CanBuildFrom[Repr, U, That]): That
def withFilter[U](p: T => Boolean): Repr.
\end{lstlisting}

@@

\smartParagraph{The lifting}
The basic idea is to use the enrich-my-library pattern to lift collection methods from \code{TraversableLike[T, Repr]} to \code{TraversableLikeOps[T, Repr]}:
\begin{lstlisting}
implicit def expToTraversableLikeOps[T, Repr](v: Exp[TraversableLike[T, Repr]]) =
  new TraversableLikeOps[T, Repr](v)
\end{lstlisting}

@@

%We use enrich-my-library again with an implicit conversion to \code{TraversableLikeOps[T, Repr]}, shown later, which corresponds to
%\code{TraversableLike[T, Repr]}.
%We provide an implicit conversion to \code{TraversableLikeOps[T, Repr]} from \code{Exp[Repr]} for any type \code{Repr} which is a subtype of \code{TraversableLike[T, Repr]} and \code{Traversable[T]} (for some element type \code{T}). This can be expressed as:
\begin{lstlisting}
implicit def expToTraversableLikeOps
  [T, Repr <: Traversable[T] with TraversableLike[T, Repr]](v: Exp[Repr]) =
    new TraversableLikeOps[T, Repr](v)
\end{lstlisting}

@@

\begin{figure}
\begin{center}
\begin{lstlisting}
class TraversableLikeOps[T, Repr <: Traversable[T] with TraversableLike[T, Repr]](t: Exp[Repr]) {
  val t: Exp[Repr]

  def withFilter(f: Exp[T] => Exp[Boolean]): Exp[Repr] = Filter(this.t, FuncExp(f))

  def map[U, That <: TraversableLike[U, That]](f: Exp[T] => Exp[U])
         (implicit c: CanBuildFrom[Repr, U, That]): Exp[That] = MapNode(this.t, FuncExp(f))
  //... other methods ...
}
//definitions of MapNode, Filter omitted.
\end{lstlisting}
\end{center}
\caption{Lifting \code{TraversableLike}}
\label{fig:liftTrav}
\end{figure}
